A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-19 paper.tex
Upadte

% Todo this is rough. Ideally we want a bit of commentary running through
% the quotations.
\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
\citet{shipilina2023origin} emphasise the reality  ``The ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events''. In contrast,
\citet{brandt2021evaluation} tie the definition of an ARG explicitly
to a generative process:
``The full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]

% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
%

The common features of these definitions is that they are expressed
in terms of evolutionary events, emphasise the exhaustiveness
of the record (full, complete), and they do not specify the mechanism
by which partial genetic inheritance occurs. Just as the graph topology
in a gARG defines the paths along which genetic inheritance occurs, some
additional information is required to define the actual regions of genome
inherited.

In the classical view of an eARG we usually have two different types
of node,
representing common ancestor and recombination events that occurred in the
history of some set of sample genomes
and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use the standard convention here
that the order of parents
is significant, but additional information can be incorporated to
distinguish left and right parents
\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.

@@

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG (left) and a corresponding
encoding of the information (right).
Recombination nodes (e.g.\ \noderef{d}) have a non-null breakpoint
and two parents.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
}
\end{figure}

>>>>

% Todo this is rough. Ideally we want a bit of commentary running through
% the quotations.
\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
\citet{shipilina2023origin} emphasise the reality  ``The ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events''. In contrast,
\citet{brandt2021evaluation} tie the definition of an ARG explicitly
to a generative process:
``The full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
In contrast ~\citet{gusfield2014recombinatorics} defines an ARG as [FIXME]
The common features of these definitions is that they are expressed
in terms of evolutionary events, and emphasise the exhaustiveness
of the record (full, complete). We will examine the consequences of both
of these properties in different contexts throughout the remainder of
this article.

Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
% % Papers that do not discuss how positions work:
% % - Mathieson and Scally: no discussion of how it works concretely (but quite
% %   close to gARG, in some ways)
% %
% The common features of these definitions is that they are expressed
% in terms of evolutionary events, emphasise the exhaustiveness
% of the record (full, complete), and they do not specify the mechanism
% by which partial genetic inheritance occurs. Just as the graph topology
% in a gARG defines the paths along which genetic inheritance occurs, some
% additional information is required to define the actual regions of genome
% inherited.
Classically, this is done by having two different types
of node:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
% In a common ancestor event the inbound lineages are merged into a
% single ancestral lineage with one parent, whereas in a recombination
% event a lineage is split into two independent
% ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use a common convention here
that the order of parents
is significant, but additional information is sometimes incorporated to
distinguish left and right parents
\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical event ARG (left) and a corresponding
encoding of the information (right).
Recombination nodes (e.g.\ \noderef{d}) have a non-null breakpoint
and two parents.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
}
\end{figure}
